Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2021-09-05 paper-revision/response-to-reviewers.tex

\reviewc{r2q4}{The authors should better emphasize the contribution from a methodological viewpoint. Indeed, without a proper emphasis on this aspect, the paper seems to be the application of an existing tool (namely, openTSNE) to a new dataset. If the contribution from a methodological viewpoint is actually limited, the authors should still put some effort in emphasizing the advantages of this approach in this specific domain.}

\answer{
We have altered the wording in ``Availability and Implementation'' section to emphasize that the extension we describe in this paper was developed by us and later incorporated into \textsf{openTSNE}. As this is the only mention of \textsf{openTSNE} in the manuscript, we believe there can be no further confusion in this respect.
}

\reviewc{r2q5}{PCA-based initialization seems to help a lot. I would appreciate to see a comparison of the results with those achievable with the simple application of PCA with just 2 principal components.}

\answer{
PCA fails to produce cluster-informative visualizations of any high-dimensional data, as it focuses on variance instead of clustering structure. Comparison of these two methods have been already and extensively published (see, for instance, ``viSNE enables visualization of high dimensional single-cell data and reveals phenotypic heterogeneity of leukemia'' supplementary material).
}

### 2021-09-05 paper-revision/diffbase.tex
Add final paper revision

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or
uniform manifold approximation and projection~\cite{umap} have recently
complemented traditional data transformation and embedding approaches such as
principal component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data from
a single coherent source, these methods may encounter problems with multiple
data sources. Here, when performing dimensionality reduction on a merged data
set, the resulting visualizations would typically reveal source-specific
clusters instead of grouping data instances of the same class, regardless of
data sources. This source-specific confounding is often referred to as {\em
domain shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift}
or {\em data set shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due to high biomedical relevance and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells, which serve as an
estimate of the expression of cell's genes.  The studies can exceed thousands
of cells and tens of thousands of genes, and typically start with cell type
analysis. Here, it is expected that cells of the same type would cluster
together in two-dimensional data visualization~\cite{seurat}. For instance,
Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
originating from the visual cortex~\cite{hrvatin2018} and the
hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also
separates the data from the two brain regions. These two regions share the same
cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
--- we would expect the data points from the two studies to overlap. Batch
effects similarly prohibit the utility of t-SNE in the exploration of
pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
a pancreatic cell atlas~\cite{baron2016} and similarly-typed cells from
diabetic patients~\cite{xin2016}. Just like with data from brain cells,
pancreatic cells cluster primarily by data source, again resulting in a
visualization driven by batch effects.

@@

The proposed solution implements a mapping of new data into an existing
t-SNE visualization. While the utility of such an algorithm was already hinted
at in recent publication~\cite{art_of_using_tsne}, we here provide its
practical and theoretically-grounded implementation. Considering the abundance
of recent publications on batch effect removal, we present surprising evidence
that a computationally more direct and principled embedding procedure solves
the batch effects problem when constructing interpretable visualizations from
different data sources.

@@

We apply the proposed approach to t-SNE visualizations of single-cell data.
Data in this realm include a variety of cells from specific tissues and are
characterized through gene expression.  In our experiments, we considered
several recently published data sets where cells were annotated with the cell
type. Our aim was to construct t-SNE visualizations where similarly-typed cells
would cluster together, despite systematic differences between data sources. To
that end, we focus on comparing different ways of using t-SNE rather than
differences to embeddings like PCA or MDS, which have been substantially
covered before~\cite{tsne,umap_single_cell}.  Below, we list the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

\begin{description}
\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
cells from the visual cortex exploring transcriptional changes after exposure
to light. This was used as a reference for the data from Chen
\etal~\cite{chen2017}, containing cells from the mouse hypothalamus and
their reaction to food deprivation. From the secondary data, we removed cells
with no corresponding types in the reference: tanycytes, ependymal,
epithelial, and unlabelled cells.

\item[Human pancreas.] Baron \etal~\cite{baron2016} created
an atlas of pancreatic cell types. We used this set as a reference
for data from Xin \etal~\cite{xin2016}, who examined transcriptional
differences between healthy and type 2 diabetic patients.

\item[Mouse retina.] Macosko \etal~\cite{macosko2015} created
an atlas of mouse retinal cell types. We used this as a reference
for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
retinal bipolar cells.
\end{description}

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after preprocessinng.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{cell_ontology}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

@@

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_pancreas.pdf}
  \caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
  cells from the same tissue from Xin \etal~\cite{xin2016}. Just like in
  Fig.~\ref{fig:transform_brain}, the vast majority of the cells from the
  secondary data set were correctly mapped to the same-typed cluster of
  reference cells.}
  \label{fig:transform_pancreas}
\end{figure}

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_retina.pdf}
  \caption{An embedding of a large reference of retinal cells from Macosco
  \etal~\cite{macosko2015} \textbf{(a)} and mapping of cells from a smaller study that
  focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} \textbf{(b)}. We use
  colors consistent with the study by Macosko \etal}
  \label{fig:transform_retina}
\end{figure}

We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
Kobak and Linderman have shown that the global consistency of embeddings produced by
popular visualization algorithms are largely dependent on their initialization~\cite{kobak2019umap}. By utilizing PCA-based initialization, t-SNE is able to achieve
more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
Another important extension is the use of multi-scale similarities, which, in addition
to considering short range interactions, also models wider point neighborhoods. Coupled
with PCA-based initializations, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
developmental origin, as discussed in Tasic \etal~\cite{tasic2018}, while this division
is less apparent when using PCA-based initialization alone in
Fig.~\ref{fig:multiscale}.b.

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbors search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; 44,808 reference, 26,830 secondary cells,
9min NN-search, 13sec optimization).

Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based 
positioning alone is insufficient. We construct a reference embedding using only
neurons from Hrvatin \etal~\cite{hrvatin2018} (Fig.~\ref{fig:optimization}.a)
and use that to position neuronal
cells from the data set from Campbell \etal~\cite{campbell2017}. We utilize the
weighted mean and median positions to initialize point positions from the secondary
data set, as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c.
After initialization, we optimize point positions using the procedure described
above for 500 iterations. The resulting visualizations from both initializations
are visually very similar, indicating stable convergence. We show one of the resulting
visualizations in Fig.~\ref{fig:optimization}.b.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{optimization.pdf}
  \caption{Comparison of different initialization schemes for positioning new data
  points onto reference embeddings. {\bf (a)} We construct a reference embedding
  using only neuronal subtypes from Hrvatin \etal~\cite{hrvatin2018}. {\bf (b)}
  We position neuronal cells from Campbell \etal~\cite{campbell2017} using the
  median initizaliation scheme from Kobak and Berens~\cite{art_of_using_tsne} and
  run optimization for 500 iterations. Compare the optimized embedding with the 
  initial median initialization {\bf (c)} or by using a simple weighted mean
  initialization {\bf (d)}.}
  \label{fig:optimization}
\end{figure}

The approach assumes that all cell types from the secondary data set are
present in the reference. The method may fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as scmap were recently proposed to cope
with such cases and identify the cells whose type is new and not included in
the reference~\cite{scmap}.

Our procedure is, therefore, asymmetrical in the choice of reference and secondary data set.
In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
Human Cell Atlas initiative~\cite{hca} make it possible to scale up our approach to wider sets of cell types.

@@

We compare our approach to two machine learning techniques, namely a $k$-nearest neighbor classifier (KNN) and a random forest ensemble, and scMap-Cluster~\cite{scmap}, a $k$-nearest neighbor based method, tailored explicitly to scRNA-seq data. scMap-Cluster uses three correlation-based distance measures and uses a voting scheme to perform classification. A threshold may also be set so that new cell types may be labeled as {\tt unrecognized}, but since our secondary data sets were chosen such that this does not happen, this was disabled. For the two machine learning approaches, we apply the typical single-cell preprocessing pipeline described in Sec.~\ref{sec:sc-pipeline}, i.e., library-size normalization, log-transformation, and select 1,000 most informative genes. Similarily to scMap-Cluster, we use the cosine distance to find the 5 nearest neighbors in the KNN model. We used 100 trees in the random forest ensemble. The models were fit on the reference data set, and no hyper-parameter tuning was performed.

@@

Due to the specific nature of single-cell data, additional steps must be taken
to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
consisting of the selection of 3,000 representative genes (see
Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
standardization, and PCA-based representation that retains 50 principal
components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply
multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to
high-dimensionality of the preprocessed input data we use cosine distance to
estimate similarities between reference data points~\cite{Domingos2012-CACM}.
When adding new data points from the secondary data set to the reference
embedding, we select 1,000 genes present in both data sets and use the cosine
similarity to estimate the similarities between the secondary data item and
reference data points. We note that similarities are computed using the raw
count matrices. The preprocessing stages are detailed in accompanying Python
notebooks (Sec.~\ref{sec:implementation}).

@@

Single-cell data sets suffer from high levels of technical noise and low
capture efficiency, resulting in sparse expression matrices~\cite{umi}. To
address this problem, we use a specialized feature-selection method, which
exploits the mean-dropout relationship of expression counts as recently
proposed by Kobak and Berens~\cite{art_of_using_tsne}. Here, genes with higher
than expected dropout rate are regarded as potential markers for cell
subpopulations and are retained in the data.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
  \caption{A comparison of standard and multi-scale t-SNE on data from the
  mouse neocortex~\cite{tasic2018}. {\bf (a)} Standard t-SNE using random 
  initialization places clusters arbitrarily. The resulting clustering structure is not globally consistent,
  as clusters of the same type of cells are dispersed throughout the landscape. Non-Neuronal clusters, for instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
  {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the clusters
  are organized in a more meaningful layout, where clusters of cells of the same type appear closer together.
  {\bf (c)} Augmenting t-SNE with multi-scale
  similarities and using proper initialization provides a more meaningful
  layout of the clusters. Non-Neuronal and Endothelial cell types are now placed in
  the same region of the embedding. There are two clear sub-groups of GABAergic
  neurons corresponding to their developmental origins, which was not as apparent
  when using clever initialization alone.}
  \label{fig:multiscale}
\end{figure}

>>>>

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{Maaten2008} or
uniform manifold approximation and projection~\cite{McInnes2018} have recently
complemented traditional data transformation and embedding approaches such as
principal component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data from
a single coherent source, these methods may encounter problems with multiple
data sources. Here, when performing dimensionality reduction on a merged data
set, the resulting visualizations would typically reveal source-specific
clusters instead of grouping data instances of the same class, regardless of
data sources. This source-specific confounding is often referred to as {\em
domain shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift}
or {\em data set shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{Butler2018,Haghverdi2018,Stuart2019}.

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due to high biomedical relevance and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells, which serve as an
estimate of the expression of cell's genes.  The studies can exceed thousands
of cells and tens of thousands of genes, and typically start with cell type
analysis. Here, it is expected that cells of the same type would cluster
together in two-dimensional data visualization~\cite{Wolf2018}. For instance,
Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
originating from the visual cortex~\cite{Hrvatin2018} and the
hypothalamus~\cite{Chen2017}. The figure reveals distinct clusters but also
separates the data from the two brain regions. These two regions share the same
cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
--- we would expect the data points from the two studies to overlap. Batch
effects similarly prohibit the utility of t-SNE in the exploration of
pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
a pancreatic cell atlas~\cite{Baron2016} and similarly-typed cells from
diabetic patients~\cite{Xin2016}. Just like with data from brain cells,
pancreatic cells cluster primarily by data source, again resulting in a
visualization driven by batch effects.

@@

The proposed solution implements a mapping of new data into an existing
t-SNE visualization. While the utility of such an algorithm was already hinted
at in recent publication~\cite{Kobak2019}, we here provide its
practical and theoretically-grounded implementation. Considering the abundance
of recent publications on batch effect removal, we present surprising evidence
that a computationally more direct and principled embedding procedure solves
the batch effects problem when constructing interpretable visualizations from
different data sources.

@@

We apply the proposed approach to t-SNE visualizations of single-cell data.
Data in this realm include a variety of cells from specific tissues and are
characterized through gene expression.  In our experiments, we considered
several recently published data sets where cells were annotated with the cell
type. Our aim was to construct t-SNE visualizations where similarly-typed cells
would cluster together, despite systematic differences between data sources. To
that end, we focus on comparing different ways of using t-SNE rather than
differences to embeddings like PCA or MDS, which have been substantially
covered before~\cite{Maaten2008,Becht2019}.  Below, we list the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

\begin{description}
\item[Mouse brain.] The data set from Hrvatin \etal~\cite{Hrvatin2018} contains
cells from the visual cortex exploring transcriptional changes after exposure
to light. This was used as a reference for the data from Chen
\etal~\cite{Chen2017}, containing cells from the mouse hypothalamus and
their reaction to food deprivation. From the secondary data, we removed cells
with no corresponding types in the reference: tanycytes, ependymal,
epithelial, and unlabelled cells.

\item[Human pancreas.] Baron \etal~\cite{Baron2016} created
an atlas of pancreatic cell types. We used this set as a reference
for data from Xin \etal~\cite{Xin2016}, who examined transcriptional
differences between healthy and type 2 diabetic patients.

\item[Mouse retina.] Macosko \etal~\cite{Macosko2015} created
an atlas of mouse retinal cell types. We used this as a reference
for the data from Shekhar \etal~\cite{Sekhar2016}, who built an atlas for
retinal bipolar cells.
\end{description}

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after preprocessinng.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{Bard2005}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

@@

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_pancreas.pdf}
  \caption{Embedding of pancreatic cells from Baron \etal~\cite{Baron2016} and
  cells from the same tissue from Xin \etal~\cite{Xin2016}. Just like in
  Fig.~\ref{fig:transform_brain}, the vast majority of the cells from the
  secondary data set were correctly mapped to the same-typed cluster of
  reference cells.}
  \label{fig:transform_pancreas}
\end{figure}

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{transform_retina.pdf}
  \caption{An embedding of a large reference of retinal cells from Macosco
  \etal~\cite{Macosko2015} \textbf{(a)} and mapping of cells from a smaller study that
  focuses on bipolar cells from Shekhar \etal~\cite{Shekhar2016} \textbf{(b)}. We use
  colors consistent with the study by Macosko \etal}
  \label{fig:transform_retina}
\end{figure}

We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
Kobak and Linderman have shown that the global consistency of embeddings produced by
popular visualization algorithms are largely dependent on their initialization~\cite{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
Another important extension is the use of multi-scale similarities, which, in addition
to considering short range interactions, also models wider point neighborhoods. Coupled
with PCA-based initializations, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
developmental origin, as discussed in Tasic \etal~\cite{Tasic2018}, while this division
is less apparent when using PCA-based initialization alone in
Fig.~\ref{fig:multiscale}.b.

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
  \caption{A comparison of standard and multi-scale t-SNE on data from the
  mouse neocortex~\cite{Tasic2018}. {\bf (a)} Standard t-SNE using random 
  initialization places clusters arbitrarily. The resulting clustering structure is not globally consistent,
  as clusters of the same type of cells are dispersed throughout the landscape. Non-Neuronal clusters, for instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
  {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the clusters
  are organized in a more meaningful layout, where clusters of cells of the same type appear closer together.
  {\bf (c)} Augmenting t-SNE with multi-scale
  similarities and using proper initialization provides a more meaningful
  layout of the clusters. Non-Neuronal and Endothelial cell types are now placed in
  the same region of the embedding. There are two clear sub-groups of GABAergic
  neurons corresponding to their developmental origins, which was not as apparent
  when using clever initialization alone.}
  \label{fig:multiscale}
\end{figure}

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{Kobak2019}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbors search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; 44,808 reference, 26,830 secondary cells,
9min NN-search, 13sec optimization).

Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based 
positioning alone is insufficient. We construct a reference embedding using only
neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref{fig:optimization}.a)
and use that to position neuronal
cells from the data set from Campbell \etal~\cite{Campbell2017}. We utilize the
weighted mean and median positions to initialize point positions from the secondary
data set, as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c.
After initialization, we optimize point positions using the procedure described
above for 500 iterations. The resulting visualizations from both initializations
are visually very similar, indicating stable convergence. We show one of the resulting
visualizations in Fig.~\ref{fig:optimization}.b.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{optimization.pdf}
  \caption{Comparison of different initialization schemes for positioning new data
  points onto reference embeddings. {\bf (a)} We construct a reference embedding
  using only neuronal subtypes from Hrvatin \etal~\cite{Hrvatin2018}. {\bf (b)}
  We position neuronal cells from Campbell \etal~\cite{Campbell2017} using the
  median initizaliation scheme from Kobak and Berens~\cite{Kobak2019} and
  run optimization for 500 iterations. Compare the optimized embedding with the 
  initial median initialization {\bf (c)} or by using a simple weighted mean
  initialization {\bf (d)}.}
  \label{fig:optimization}
\end{figure}

The approach assumes that all cell types from the secondary data set are
present in the reference. The method may fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as scmap were recently proposed to cope
with such cases and identify the cells whose type is new and not included in
the reference~\cite{Kiselev2018}.

Our procedure is, therefore, asymmetrical in the choice of reference and secondary data set.
In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
Human Cell Atlas initiative~\cite{Rozenblatt-Rosen2017} make it possible to scale up our approach to wider sets of cell types.

@@

We compare our approach to two machine learning techniques, namely a $k$-nearest neighbor classifier (KNN) and a random forest ensemble, and scMap-Cluster~\cite{Kiselev2018}, a $k$-nearest neighbor based method, tailored explicitly to scRNA-seq data. scMap-Cluster uses three correlation-based distance measures and uses a voting scheme to perform classification. A threshold may also be set so that new cell types may be labeled as {\tt unrecognized}, but since our secondary data sets were chosen such that this does not happen, this was disabled. For the two machine learning approaches, we apply the typical single-cell preprocessing pipeline described in Sec.~\ref{sec:sc-pipeline}, i.e., library-size normalization, log-transformation, and select 1,000 most informative genes. Similarily to scMap-Cluster, we use the cosine distance to find the 5 nearest neighbors in the KNN model. We used 100 trees in the random forest ensemble. The models were fit on the reference data set, and no hyper-parameter tuning was performed.

@@

Due to the specific nature of single-cell data, additional steps must be taken
to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
consisting of the selection of 3,000 representative genes (see
Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
standardization, and PCA-based representation that retains 50 principal
components~\cite{Stuart2019,Wolf2018}. To obtain the reference embedding, we apply
multi-scale t-SNE using PCA initialization~\cite{Kobak2019}. Due to
high-dimensionality of the preprocessed input data we use cosine distance to
estimate similarities between reference data points~\cite{Domingos2012}.
When adding new data points from the secondary data set to the reference
embedding, we select 1,000 genes present in both data sets and use the cosine
similarity to estimate the similarities between the secondary data item and
reference data points. We note that similarities are computed using the raw
count matrices. The preprocessing stages are detailed in accompanying Python
notebooks (Sec.~\ref{sec:implementation}).

@@

Single-cell data sets suffer from high levels of technical noise and low
capture efficiency, resulting in sparse expression matrices~\cite{umi}. To
address this problem, we use a specialized feature-selection method, which
exploits the mean-dropout relationship of expression counts as recently
proposed by Kobak and Berens~\cite{Kobak2019}. Here, genes with higher
than expected dropout rate are regarded as potential markers for cell
subpopulations and are retained in the data.

@@
